Creating figures for the paper

### 2024-08-08 paper/reports/paper_figures.Rmd
move paper prep to separate subdir, integrate Matt's plotting functions

>>>>

```{r}
# dirs
lh22_dir <- "/lustre/scratch125/casm/team268im/lh22/Rheumatoid_Arthritis_Paper/"
nb15_dir <- "/lustre/scratch126/casm/team154pc/nb15/liver/"

# functions and libraries
source("bin/lolliplot.R")
library(magrittr)
library(ggplot2)

# load the annotated mutations from nanoseq output
load(file.path(lh22_dir, "Figures/Figure_2/LCM_NanoSeq_Overlap/GOI",
               "pileup_muts.rda"))
nano_muts <- nano_muts %>% tibble::as_tibble()

# run dNdScv
dndscv_in <- nano_muts %>% dplyr::select(sampleID, chr, pos, ref, mut)
dndscv_out <- dndscv::dndscv(dndscv_in)

# load refcds, add gene names
data("refcds_hg19", package = "dndscv")
names(RefCDS) <- RefCDS %>% purrr::map(~ .$gene_name)

# combine annotated mutations with dnds output
nanoseq_data <-
  nano_muts %>%
  dplyr::mutate(impact = dplyr::case_when(impact == "no-SNV" ~ "Indel",
                                          is.na(impact) ~ "Non-coding",
                                          TRUE ~ impact)) %>%
  dplyr::select(sampleID, chr, pos, ref, mut, duplex_cov, impact, donor) %>%
  dplyr::left_join(dndscv_out$annotmuts) %>%
  # rename cols
  dplyr::rename(CHROM = chr, POS = pos, REF = ref, ALT = mut)

# plotting params
filterNonCoding <- TRUE
pointWidth <- 0.017

# define genes of interest
g <- c("NF1")

# get gene info
g_CDS <- RefCDS[[g]]

#  get gene intervals
intervals_cds0 <- tibble::tibble(
  chr = g_CDS$chr,
  strand = g_CDS$strand,
  start = g_CDS$intervals_cds[, 1],
  end = g_CDS$intervals_cds[, 2]
) %>%
  dplyr::arrange(start) %>%
  dplyr::mutate(length = (end + 1) - start)

# get genomic position to CDS position index
cds_coords_index <- getCDSindex(intervals_cds0)

# convert CDS intervals
intervals_cds <-
  intervals_cds0 %>%
  dplyr::left_join(cds_coords_index %>%
                     dplyr::select(start = pos, cdsStart = cdsPos),
                   by = "start") %>%
  dplyr::left_join(cds_coords_index %>%
                     dplyr::select(end = pos, cdsEnd = cdsPos),
                   by = "end")

# subset nanoseq_data to variants in the gene
gene0 <-
  nanoseq_data %>%
  dplyr::filter(CHROM == unique(intervals_cds$chr),
                POS >= min(intervals_cds$start),
                POS <= max(intervals_cds$end))

# get coords outside the CDS
noncds <-
  cds_coords_index %>%
  dplyr::rename(POS = pos) %>%
  dplyr::bind_rows(getCDSindexNonCDS(cds_coords_index, gene0,
                                     strand = unique(intervals_cds$strand)))

gene <-
  gene0 %>%
  dplyr::left_join(noncds, by = "POS") %>%
  dplyr::arrange(cdsPos) %>%
  tidyr::drop_na(cdsPos)

# set impact colours
impact_palette <-
  c("Non-coding" = "gray100",
    "Synonymous" = "gray76",
    "Missense" = "#abd9e9",
    "Nonsense" = "#fdae61",
    "Indel" = "pink",
    "Essential_Splice" = "#fee090",
    "Stop_loss" = "#bb0f12")
gene <-
  gene %>%
  dplyr::mutate(impact = factor(impact, levels = names(impact_palette)),
                col = impact_palette[impact])

# set label positions
lab_pos <-
  jitterLabels(sort(gene$cdsPos),
               xscale = c(1, max(cds_coords_index$cdsPos)),
               lineW = pointWidth) %>%
  reAdjustLabels(lineW = pointWidth, xmax = max(cds_coords_index$cdsPos))
variants <-
  gene %>%
  dplyr::mutate(labPos = lab_pos,
                aachange = tidyr::replace_na(aachange, "."))

# xmax and xmin
x_max <-
  max(c(intervals_cds$cdsEnd, intervals_cds$cdsStart)) +
  max(c(intervals_cds$cdsEnd, intervals_cds$cdsStart)) * 0.05
x_min <- 1 - x_max * 0.01

# plot variants
p_var <-
  variants %>%
  dplyr::group_by(CHROM, POS, REF, ALT) %>%
  dplyr::mutate(n_donors = dplyr::n_distinct(donor)) %>%
  ggplot() +
  # variant stems
  geom_segment(aes(x = cdsPos, xend = cdsPos, y = 0,
                   yend = (0.125 * max(lab_pos)))) +
  geom_segment(aes(x = cdsPos, xend = labPos, y = (0.125 * max(lab_pos)),
                   yend = (0.25 * max(lab_pos)))) +
  geom_segment(aes(x = labPos, xend = labPos, y = (0.25 * max(lab_pos)),
                   yend = (0.325 * max(lab_pos)))) +
  # variant points
  geom_point(aes(x = labPos, y = 0.325 * max(lab_pos), fill = impact,
                 size = n_donors),
             shape = 21, colour = "black", size = 5) +
  # label aa change
  # geom_text(aes(label = name, x = labPos, y = (0.5 * max(lab_pos))),
  #           angle = 90) +
  theme_void() +
  scale_fill_manual(guide = "legend", values = impact_palette) +
  scale_x_continuous(expand = c(0, 0), limits = c(x_min, x_max)) +
  coord_cartesian(clip = "off") +
  scale_y_continuous(limits = c(0, max(lab_pos)), expand = c(0, 0)) +
  ggtitle(g)

# plot CDS
p_cds <-
  intervals_cds %>%
  ggplot(aes(xmin = cdsStart, xmax = cdsEnd, ymin = 0, ymax = 1)) +
  geom_rect(fill = "#d0d1e6", colour = "black") +
  theme_void() + 
  xlab("CDS") +
  scale_x_continuous(expand = c(0, 0), limits = c(x_min, x_max)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme(plot.margin = unit(c(0, 0.5, 0.5, 0.5), "cm"))

pdf("test.pdf", width = 20)
cowplot::plot_grid(p_var, p_cds, ncol = 1, rel_heights = c(15, 1))
dev.off()

```
